A data-driven, pixel-level model to improve the precision of \Kepler\ photometry

### 2013-10-18 documents/plm.tex
tweaks, careful about words model and predict

\begin{abstract}
The precision of \Kepler\ photometry%
  ---the most precise photometric measurements of stars ever made---%
  appears to be limited by unknown or untracked variations
  in spacecraft pointing, point-spread function, stellar motions and parallaxes, and temperature.
Here we present \name,
  a data-driven model of these variations in \Kepler\ data (or really of their impact on the photometry).
Importantly (and uniquely),
  \name\ works at the pixel level (not the photometric measurement level);
  it can capture more fine-grained information about the variation of the spacecraft
  (especially regarding pointing and point-spread function)
  than is available in the pixel-summed photometry.
\name\ is extreme in that it has enormous flexibility and it only uses data (and meta-data) to model data:
\name\ provides a prediction for each readout pixel built from a linear combination of the readouts
  from very large numbers of nearby pixels (plus some spacecraft information);
  the choice of nearby pixels is guided by ideas from causal inference.
\name\ avoids over-fitting by employing a train-and-test formalism
  designed to ensure that transit-like photometric events and short-timescale stellar variability
  cannot be captured or distorted by the model;
  it is designed to remove spacecraft-induced variability but not intrinsic stellar variability.
We show that \name\ outperforms the \Kepler\ Presearch (\sic) Data Conditioning method on a set of example stars.
We release open-source code that provides \name\ output for any star in the existing \Kepler\ Archive,
  and we discuss applications for other missions,
  including any possible \Kepler\ two-wheel mission.
\end{abstract}

@@

One such method---one that is designed to model spacecraft-induced problems
  but \emph{not} interfere with measurements of stellar variability---%
  is the \Kepler\ Presearch (\sic) Data Conditioning (PDC) method (\todo{citation}).
The PDC ``de-trends'' the \Kepler\ lightcurves by fitting them with a small set of basis lightcurves
  generated from a principal components analysis (PCA) of filtered lightcurves.
That is, it uses data to model data,
  regularizing the fit (and avoiding over-fitting) by filtering and restricting the dimensionality (through PCA).
The method proposed here, \name, is very similar in spirit to the PDC.
The main differences are
  (1)~that \name\ works in the pixel domain, not the lightcurve domain, so it has access to more fine-grained information,
  (2)~that \name\ has far more freedom (far more parameters) than the PDC
  but it strictly avoids over-fitting the lightcurves on exoplanet-transit time-scales
  through strong regularization and a train-and-test framework, and
  (3)~that when \name\ fits a star, there is no possibility of any contribution of the star itself to the fit basis vectors.
The two methods (\name\ and PDC) are similar however,
  in that they both make the assumption that whatever spacecraft effects are imprinting variability on a stellar lightcurve
  must also be imprinting similar or related variability on other lightcurves.
They also assume that the relationships can be captured with linear models.
By going to the pixel level (unlike the PDC, which works at the photometry level),
  \name\ makes it easier for the model to capture variability
  that is coming through variations in the centroids and point-spread function
  from spacecraft pointing, roll, and temperature.

@@

The first general idea about data-driven modeling of the kind used here
  is that each data point or data source is going to be modeled
  (or really \emph{predicted})
  with a parameterized mathematical function of \emph{other data}.
That is, given a choice of some \emph{target} \Kepler\ data,
  we are going to find parameters of a function that takes as input some of the other \Kepler\ data,
  and provides as output predictions of the target data.
The simplest models are \emph{linear models},
  in which predictions for the target data are built from linear combinations of the other data,
  and in which the objective functions are quadratic in the prediction residual.
Examples of quadratic objective functions include Gaussian likelihoods, mean-squared-error, and chi-square statistics.
These models are simple,
  not just because they are easy to express and compute,
  but because optimization is convex:
There is only one optimum for the objective function.

@@

The third general idea is that we need to control for over-fitting.
That is, once you have a flexible-enough model you can in principle fit \emph{anything},
  whether it was caused by the spacecraft, intrinsic stellar variability, or a transiting exoplanet.
How do you prevent a very flexible model from taking small noise-induced fluctuations in all the input data
  and carefully combining them linearly into detailed models for every nook and cranny in the target data?
In most projects in astrophysics, over-fitting is controlled for by limiting model freedom.
The model is restricted in the number of parameters
  (as in ``you can't have more parameters than data points'')
  or by limiting the dimensionality
  (as in principal components analysis)
  or by applying strong priors
  (as with smoothness priors, that effectively reduce freedom without explicitly reducing the number of parameters).
In each case, the restriction on the model is controlled by some \emph{hyper-parameter}
  (such as the number of inputs, or number of principal components, or the strength of the smoothness prior).
The hyper-parameters can be set or tested with tools like cross-validation,
  the fully marginalized likelihood (Bayes factor or evidence),
  chi-squared statistical criteria,
  or intuition or heuristics.
Here we take a different and more general approach, which is to use a train-and-test framework.

@@

In the context of this project, the quantities of interest are the light-curves
  (photometry as a function of time or apparent flux as a function of time)
  for every star.
What we get, for every star, from the data-driven model, is a train-and-test prediction
  for the brightness of every pixel in the patch of the detector relevant to the star,
  along with the residual, or the difference between the measured read-out pixel value and the prediction (data minus model).
What will be called the ``\name\ photometry'' in what follows is constructed from what will be called the ``\name\ images''.
The \name\ photometry is constructed from the \name\ images
  just as the \Kepler\ official SAP photometry is constructed from the \todo{XXX images};
  that is, by summing pre-defined sets of pixels (with unit weights; see above and below for criticism of this).
The \name\ images are constructed from the data-driven model predictions by taking,
  for each pixel in each contiguous data segment (see below),
  the \emph{mean} predicted pixel value (mean over time),
  and adding to it the residuals (data minus model) away from the data-driven model prediction.
These \name\ images are ``calibrated'' images,
  in that the data-driven model has been used to fit out all the variability it can,
  subject to the restrictions imposed by its finite freedom and the train-and-test specification.
The signal---in which exoplanet transits will be found---is fundamentally the observed residual away from this prediction,
  but it has to be added back into the mean prediction to have the right ``baseline'' or zero level.
In short, \name\ will produce prediction images, residual images, \name\ images (mean prediction plus residual),
  and aperture-based \name\ photometry analogous to the SAP photometry.

@@

We specify what likelihood and what regularization?
\todo{foo:}
\begin{eqnarray}
I_{mn}         &=& I^{\ast}_{mn} + \mbox{noise}
\\
I^{\ast}_{mn}  &=& \sum_{m'\in\set{M}_m} a_{mm'}\,I_{m'n}
\\
I^{\ast}_{mn'} &=& \sum_{m'\in\set{M}_m} a_{mm'}\,I_{m'n'}
\quad,
\end{eqnarray}
where \todo{bar}
\begin{eqnarray}
\chi^2_{mn}    &=& \sum_{n'\in\set{N}_n} \frac{[I_{mn'} - I^{\ast}_{mn'}]^2}{\sigma^2_{mn'}}
                 + \sum_{m'\in\set{M}_m} \frac{a_{mm'}^2}{\Sigma^2}
\quad,
\end{eqnarray}
where \todo{bar}

>>>>

\begin{abstract}
The precision of \Kepler\ photometry%
  ---the most precise photometric measurements of stars ever made---%
  appears to be limited by unknown or untracked variations
  in spacecraft pointing, point-spread function, stellar motions and parallaxes, and temperature.
Here we present \name,
  a data-driven model of these variations in \Kepler\ data (or really of their impact on the photometry).
Importantly (and uniquely),
  \name\ works at the pixel level (not the photometric measurement level);
  it can capture more fine-grained information about the variation of the spacecraft
  (especially regarding pointing and point-spread function)
  than is available in the pixel-summed photometry.
\name\ is extreme in that it has enormous flexibility and it only uses data (and meta-data) to predict data:
\name\ provides a prediction for each readout pixel built from a linear combination of the readouts
  from very large numbers of nearby pixels (plus some spacecraft information);
  the choice of nearby pixels is guided by ideas from causal inference.
\name\ avoids over-fitting by employing a train-and-test formalism
  designed to ensure that transit-like photometric events and short-timescale stellar variability
  cannot be captured or distorted by the model;
  it is designed to remove spacecraft-induced variability but not intrinsic stellar variability.
We show that \name\ outperforms the \Kepler\ Presearch (\sic) Data Conditioning method on a set of example stars.
We release open-source code that provides \name\ output for any star in the existing \Kepler\ Archive,
  and we discuss applications for other missions,
  including any possible \Kepler\ two-wheel mission.
\end{abstract}

@@

One such method---one that is designed to describe or model spacecraft-induced problems
  but \emph{not} interfere with measurements of stellar variability---%
  is the \Kepler\ Presearch (\sic) Data Conditioning (PDC) method (\todo{citation}).
The PDC ``de-trends'' the \Kepler\ lightcurves by fitting them with a small set of basis lightcurves
  generated from a principal components analysis (PCA) of filtered lightcurves.
That is, it uses data to predict data,
  regularizing the fit (and avoiding over-fitting) by filtering and restricting the dimensionality (through PCA).
The method proposed here, \name, is very similar in spirit to the PDC.
The main differences are
  (1)~that \name\ works in the pixel domain, not the lightcurve domain, so it has access to more fine-grained information,
  (2)~that \name\ has far more freedom (far more parameters) than the PDC
  but it strictly avoids over-fitting the lightcurves on exoplanet-transit time-scales
  through strong regularization and a train-and-test framework, and
  (3)~that when \name\ fits a star, there is no possibility of any contribution of the star itself to the fit basis vectors.
The two methods (\name\ and PDC) are similar however,
  in that they both make the assumption that whatever spacecraft effects are imprinting variability on a stellar lightcurve
  must also be imprinting similar or related variability on other lightcurves.
They also assume that the relationships can be captured with linear models.
By going to the pixel level (unlike the PDC, which works at the photometry level),
  \name\ makes it easier for the model to capture variability
  that is coming through variations in the centroids and point-spread function
  from spacecraft pointing, roll, and temperature.

@@

The first general idea about data-driven modeling of the kind used here
  is that each data point or data source is going to be 
  \emph{predicted} with or using a parameterized mathematical function of \emph{other data}.
That is, given a choice of some \emph{target} \Kepler\ data,
  we are going to find parameters of a function that takes as input some of the other \Kepler\ data,
  and provides as output predictions of the target data.
The simplest models are \emph{linear models},
  in which predictions for the target data are built from linear combinations of the other data,
  and in which the objective functions are quadratic in the prediction residual.
Examples of quadratic objective functions include Gaussian likelihoods, mean-squared-error, and chi-square statistics.
These models are simple,
  not just because they are easy to express and compute,
  but because optimization is convex:
There is only one optimum for the objective function.

@@

The third general idea is that we need to control for over-fitting.
That is, once you have a flexible-enough model you can in principle fit \emph{anything},
  whether it was caused by the spacecraft, intrinsic stellar variability, or a transiting exoplanet.
How do you prevent a very flexible model from taking small noise-induced fluctuations in all the input data
  and carefully combining them linearly into detailed models for every nook and cranny in the target data?
In most projects in astrophysics, over-fitting is controlled for by limiting model freedom.
The model is restricted in the number of parameters
  (as in ``you can't have more parameters than data points'')
  or by limiting the dimensionality
  (as in principal components analysis)
  or by applying strong priors
  (as with smoothness priors, that effectively reduce freedom without explicitly reducing the number of parameters).
In each case, the restriction on model freedom is controlled by some \emph{hyper-parameter}
  (such as the number of inputs, or number of principal components, or the strength of the smoothness prior).
The hyper-parameters can be set or tested with tools like cross-validation,
  the fully marginalized likelihood (Bayes factor or evidence),
  chi-squared statistical criteria,
  or intuition or heuristics.
Here we take a different and more general approach, which is to use a train-and-test framework.

@@

In the context of this project, the quantities of interest are the light-curves
  (photometry as a function of time or apparent flux as a function of time)
  for every star.
What we get, for every star, from the data-driven model, is a train-and-test prediction
  for the brightness of every pixel in the patch of the detector relevant to the star,
  along with the residual, or the difference between the measured read-out pixel value and the prediction (data minus prediction).
What will be called the ``\name\ photometry'' in what follows is constructed from what will be called the ``\name\ images''.
The \name\ photometry is constructed from the \name\ images
  just as the \Kepler\ official SAP photometry is constructed from the \todo{XXX images};
  that is, by summing pre-defined sets of pixels (with unit weights; see above and below for criticism of this).
The \name\ images are constructed from the data-driven model predictions by taking,
  for each pixel in each contiguous data segment (see below),
  the \emph{mean} predicted pixel value (mean over time),
  and adding to it the residuals (data minus prediction) away from the data-driven model prediction.
These \name\ images are ``calibrated'' images,
  in that the data-driven model has been used to fit out all the variability it can,
  subject to the restrictions imposed by its finite freedom and the train-and-test specification.
The signal---in which exoplanet transits will be found---is fundamentally the observed residual away from this prediction,
  but it has to be added back into the mean prediction to have the right ``baseline'' or zero level.
In short, \name\ will produce prediction images, residual images, \name\ images (mean prediction plus residual),
  and aperture-based \name\ photometry analogous to the SAP photometry.

@@

The model is
We specify what likelihood and what regularization?
\todo{foo:}
\begin{eqnarray}
I_{mn}         &=& I^{\ast}_{mn} + \mbox{noise}
\\
I^{\ast}_{mn}  &=& \sum_{m'\in\set{M}_m} a_{mm'}\,I_{m'n}
\quad,
\end{eqnarray}
where \todo{bar}
\begin{eqnarray}
\chi^2_{mn}    &=& \sum_{n'\in\set{N}_n} \frac{[I_{mn'} - I^{\ast}_{mn'}]^2}{\sigma^2_{mn'}}
                 + \sum_{m'\in\set{M}_m} \frac{a_{mm'}^2}{\Sigma^2}
\\
I^{\ast}_{mn'} &=& \sum_{m'\in\set{M}_m} a_{mm'}\,I_{m'n'}
\quad,
\end{eqnarray}
where \todo{bar}
